Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-10-18 paper/ms.tex
More work on the introduction.

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be overcome in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
goals:  1). to analyze the frequency spectrum of stellar variability to beter understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism 
\citep{Rasmussen, Gibson}.  The basis of this method is to assume that stellar variability
is random, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the 
correlation matrix is modeled with a time-independent autocorrelation function (although 
the time-independence can be relaxed), including a white-noise component that may vary
with time, and the autocorrelation function is described by a functional form that may be 
simply parameterized by a relatively small number of parameters, sometimes referred to as 
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.

@@

\begin{enumerate}
\item Non-parametric modeling of time series with correlated noise using Gaussian processes
\item Assumptions:\\
  a). Stationary noise (although this can be broken).\\
  b). Gaussian
\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
   curves; Solar curves; short-cadence.
\item Solution(s):\\
  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
\item Problem: HODLR is tricky to use; CARMA requires certain relations between
  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
   $O(N*p^2)$, where p is number of Lorentzians.
  - Sum of Lorentzians is a good description of stellar variability:
     - granulation has expoential (or sums of exponential) noise
     - asteroseismic obeys sums of Lorentzians
     - periodic variability due to spots/plages/etc. can also be
       approximated by Lorentzian(s)
\item Generalized Ambikasarn for complex coefficients (Hermitian).
\item Performance:\\
  a). Show $O(N*p^2)$ scaling\\
  b). Show comparison to HODLR.\\
  c). Show comparison to wavelet (ala Carter \& Winn).\\
  d). Compare with CARMA Kallman filter(?).
\item Various methods/formulae:\\
  a). Likelihood computation.\\
  b). Generating GP with particular covariance [yet to solve]. [ ]\\
  c). Inferring particular components.\\
  d). Computing derivatives of GPs (as used in RV methods).\\
  e). Forecasting/interpolating.\\
  f). Inferring confidence intervals.\\
  g). Derivative of likelihood function with respect to kernel parameters.\\
  h). Non-stationary GP (can the coefficients vary with time?).\\
  i). Multi-band (each element of extended matrix becomes a matrix).
  j). Approximating various kernels: Matern, exp(sin^2), Gaussian, etc.
\item Generalization to non-stationary noise.
\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Analysis of solar data.
  c). CO2 Mauna Kea
\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.\\
  e). Better calibration of the flicker log(g) relation.\\
  f). Measurement of rotational periods - gyrochronology.\\
  g). Doppler shifts in oscillating EBs.

>>>>

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be overcome in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to beter understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism 
\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the 
correlation matrix is modeled with a time-independent autocorrelation function (although 
the time-independence can be relaxed), including a white-noise component that may vary
with time, and the autocorrelation function is described by a functional form that may be 
simply parameterized by a relatively small number of parameters, sometimes referred to as 
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
The goal of planet detection includes measuring the Doppler shift of the star, or measuring
the decrement of flux as a planet transits a star.  To properly account for the impact of 
stellar variability on the planet parameters requires treating correlations in the data, and 
to do so also requires evaluating the correlation matrix.

The size of the correlation matrix, $N_{elements}$ scales as the square of the length of the time-series,
$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become 
prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
using long time series datsets, such as the \texttt{Kepler} dataset $N_{time} \approx 10^{5-6}$,
\texttt{Spitzer} time series with $N_{time} \approx 10^5$, and solar time series with
$N_{time} \approx 10^7$.  So faster means of solving for the likelihood is required in these
cases.

One promising solution is to use approximate techiques based upon the fact that correlation
matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the 
solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
solver is complex and can be challenging to use with general choices of kernel.  Another
promising approximate technique is to use a uniform resampling of the dataset, and then
rely on the structure of Toeplitz matrices, which require uniformly space time series,
to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
this approach has the drawback of being approximate.  A third approach is to use a kernel
based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
of the Lorentzians are required to have a specific relation.  In addition, these
techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

The approach we explore in this paper is based upon a method developed by Press \&
Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
They observed how a combination of two exponential kernels would have a simliar scaling,
but they did not demonstrate how to extend this to more than two kernel components.

A significant advance was made by generalizing the Press-Rybicki method to an
arbitrary number of exponential kernels by utilizing the fact that the elements of
such matrices can be written in terms of the products of components of two vectors,
which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
Rybicki-Press' (\texttt{GenRP}) approach
has many advantages:  the computation time and storage both scale as $O(N_{time})$,
the coefficients of the kernels need not be stationary, the computation of the
likelihood is exact, and the algorithm is simple to implement, and thus is robust.
Unlike the \texttt{CARMA} model, there is no restriction on the relation between
the coefficients (except that the kernel needs to be positive definite).  Unlike
\texttt{HODLR} and \texttt{KISS-GP} the method is exact.  The main drawback of
this approach is that the sum of exponential kernels (with real coefficients) can 
only represent a simplified range power spectra of auto-correlated time series
that can be expressed the sum of Lorentzians that peak at zero frequency.

However, this drawback turns out to not be significant:  the exponential kernel
can have a complex exponential coefficients (plus their complex-conjugates), and 
thus approximate an arbitrary autocorrelation function with enough terms, including
quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
a {\it very} good approximation of stellar variability spectra:  the granulation
noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
approach to kernels of the form:
\begin{equation}
K(\tau) = a_0 \delta(\tau) + \sum_j a_{j=1}^{N_{kernel}} e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
and zero otherwise.  We show that
the extended matrix that embeds this multi-component kernel can be written
in real notation, which speeds up the evaluation of the likelihood function,
and allows optimization software to use automatic differential in computing
derivatives of the likelihood function.

The plan of the paper is...

@@

\begin{enumerate}
%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
%\item Assumptions:\\
%  a). Stationary noise (although this can be broken).\\
%  b). Gaussian
%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
%   curves; Solar curves; short-cadence.
%\item Solution(s):\\
%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
%  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
%  d). KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp. 1775-1784, 2015
%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
%\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
%   $O(N*p^2)$, where p is number of Lorentzians.
%  - Sum of Lorentzians is a good description of stellar variability:
%     - granulation has expoential (or sums of exponential) noise
%     - asteroseismic obeys sums of Lorentzians
%     - periodic variability due to spots/plages/etc. can also be
%       approximated by Lorentzian(s)
%\item Generalized Ambikasarn for complex coefficients (Hermitian).
\item Performance:\\
  a). Show $O(N*p^2)$ scaling\\
  b). Show comparison to HODLR.\\
  c). Show comparison to wavelet (ala Carter \& Winn).\\
  d). Compare with CARMA Kallman filter(?).
\item Various methods/formulae:\\
  a). Likelihood computation.\\
  b). Generating GP with particular covariance [yet to solve]. [ ]\\
  c). Inferring particular components.\\
  d). Computing derivatives of GPs (as used in RV methods).\\
  e). Forecasting/interpolating.\\
  f). Inferring confidence intervals.\\
  g). Derivative of likelihood function with respect to kernel parameters.\\
  h). Non-stationary GP (can the coefficients vary with time?).\\
  i). Multi-band (each element of extended matrix becomes a matrix).
  j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
\item Generalization to non-stationary noise.
\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Analysis of solar data.
  c). CO2 Mauna Kea
  d). TBD Computing derivatives of GPs (as used in RV methods).\\
  e). TBD Forecasting/interpolating.\\
  f). TBD Inferring confidence intervals.\\
  g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
  h). TBD Non-stationary GP (can the coefficients vary with time?).\\
  i). TBD Multi-band (each element of extended matrix becomes a matrix).

\item Example application: TYC 3559\\
  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
  b). Simulated datasets - hyperparameter/PSD inference.\\
  c). Analysis of solar data?

\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.\\
  e). Better calibration of the flicker log(g) relation.\\
  f). Measurement of rotational periods - gyrochronology.\\
  g). Doppler shifts in oscillating EBs.
